---
title: "Dude, Where's My Parking Requirement? Chicago Says You Don't Need One"
author: "Michael McLean, Member of Abundant Housing Illinois"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: hide
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  warning = FALSE, 
  message = FALSE,
  fig.width = 10,
  fig.height = 8
)
```

## Executive Summary

On July 16, 2025, Chicago City Council [passed](https://chi.streetsblog.org/2025/07/18/chicagos-vote-to-totally-eliminate-parking-mandates-near-train-stations-is-great-news-creating-transit-friendly-housing) a landmark ordinance (O2025-0015577) that dramatically expands areas where residential developments can proceed without parking mandates. This map visualizes the new landscape of parking-free development zones across the city.

**Key Changes:**
- Properties within CTA-served areas (outside downtown) can now build without parking
- Two exceptions remain: Metra-only areas and downtown zoning districts still require Administrative Adjustments

**Methodology Caveat:** The Metra station buffers shown in this analysis are calculated from GTFS station centerpoints, not from actual station entrances/exits. In practice, the 2,640-foot radius should be measured from platform access points, which may result in slightly different coverage areas than shown here.


<div class="alert alert-info">
**Dear Reader:** If this kind of policy excites you, please consider becoming a member of **[Abundant Housing Illinois](https://abundanthousingillinois.org/)**, which will get you access to the YIMBY Action Slack where over 400 members discuss and organize for housing abundance.
</div>


```{r packages, include=FALSE}
## Setup and Dependencies

# Load required packages
packages <- c("tidyverse", "sf", "leaflet", "leaflet.extras", 
              "httr", "jsonlite", "data.table", "units", "glue")

# Install missing packages
new_packages <- packages[!packages %in% installed.packages()[,"Package"]]
if (length(new_packages) > 0) {
  install.packages(new_packages)
}

# Load libraries
invisible(lapply(packages, library, character.only = TRUE))

# Set consistent CRS codes
WGS84 <- 4326  # Web mapping standard
IL_STATE_PLANE <- 3435  # Illinois State Plane (feet) for accurate buffers

sf_use_s2(FALSE)
```


```{r helper_functions, include=FALSE}

## Data Collection Functions

# First, let's define helper functions to handle Chicago's data portal and GTFS feeds:

#' Fetch complete dataset from Chicago's Socrata portal
#' @param dataset_id Socrata dataset identifier
#' @param chunk_size Maximum rows per API call (default: 50000)
#' @return sf object with all features
fetch_socrata_geojson <- function(dataset_id, chunk_size = 50000) {
  base_url <- glue("https://data.cityofchicago.org/resource/{dataset_id}.geojson")
  offset <- 0
  all_features <- list()
  
  # Loop through dataset in chunks
  repeat {
    query_url <- glue("{base_url}?$limit={chunk_size}&$offset={offset}")
    
    # Attempt to read chunk
    chunk <- tryCatch(
      st_read(query_url, quiet = TRUE),
      error = function(e) NULL
    )
    
    # Check if we've reached the end
    if (is.null(chunk) || nrow(chunk) == 0) break
    
    all_features[[length(all_features) + 1]] <- chunk
    
    # If chunk is smaller than limit, we've got everything
    if (nrow(chunk) < chunk_size) break
    
    offset <- offset + chunk_size
  }
  
  # Combine all chunks
  combined <- do.call(rbind, all_features)
  return(st_make_valid(combined))
}

#' Download and cache files locally
#' @param url Download URL
#' @param filename Local filename
#' @return Path to local file
download_and_cache <- function(url, filename) {
  if (!file.exists(filename)) {
    download.file(url, filename, mode = "wb", quiet = TRUE)
  }
  return(filename)
}

#' Extract GTFS zip file
#' @param zip_path Path to GTFS zip
#' @param extract_dir Directory name for extraction
#' @return Path to extracted directory
extract_gtfs <- function(zip_path, extract_dir) {
  if (!dir.exists(extract_dir)) {
    unzip(zip_path, exdir = extract_dir)
  }
  return(extract_dir)
}

#' Query ArcGIS REST service
#' @param layer_id Layer ID from Chicago's zoning service
#' @return sf object with features
query_arcgis <- function(layer_id) {
  base <- "https://gisapps.chicago.gov/arcgis/rest/services/ExternalApps/Zoning/MapServer"
  query_url <- glue("{base}/{layer_id}/query?where=1%3D1&outFields=*&outSR=4326&f=geojson")
  return(st_read(query_url, quiet = TRUE))
}
```

```{r load_zoning, include=FALSE}

## Step 1: Load Zoning Districts

# We begin by loading Chicago's zoning polygons and identifying downtown "D" districts, which remain subject to parking requirements:

# Fetch all zoning polygons (~30,000 features)
cat("Loading Chicago zoning districts...\n")
zoning_sf <- fetch_socrata_geojson("dj47-wfun")

# Extract downtown districts (those starting with "D")
downtown_districts <- zoning_sf %>%
  filter(str_starts(zone_class, "D")) %>%
  st_make_valid()

cat(glue("Found {nrow(downtown_districts)} downtown zoning polygons\n"))
```


```{r gtfs_processing, include=FALSE}
## Step 2: Process Transit Station Data

# Next, we'll download and process CTA and Metra GTFS data to identify rail stations:

# Download GTFS feeds
cat("Downloading transit schedules...\n")
cta_zip <- download_and_cache(
  "https://www.transitchicago.com/downloads/sch_data/google_transit.zip", 
  "cta_gtfs.zip"
)

metra_zip <- download_and_cache(
  "https://schedules.metrarail.com/gtfs/schedule.zip", 
  "metra_gtfs.zip"
)

# Extract files
cta_dir <- extract_gtfs(cta_zip, "cta")
metra_dir <- extract_gtfs(metra_zip, "metra")

# Load stop data
cta_stops <- fread(file.path(cta_dir, "stops.txt"))
metra_stops <- fread(file.path(metra_dir, "stops.txt"))
```

```{r identify_rail_stops, include=FALSE}
# Identify CTA rail stations (parent stations or location_type = 1)
cta_rail_stops <- cta_stops %>%
  filter(
    (!is.na(parent_station) & parent_station != "") |
    (!is.na(location_type) & location_type == 1)
  )

# Fallback method if no parent stations found
if (nrow(cta_rail_stops) == 0) {
  cat("Using route-based method to identify CTA rail stops...\n")
  
  cta_routes <- fread(file.path(cta_dir, "routes.txt"))
  rail_routes <- cta_routes %>% filter(route_type == 1)  # 1 = subway/metro
  
  cta_trips <- fread(file.path(cta_dir, "trips.txt"))
  rail_trips <- cta_trips %>% filter(route_id %in% rail_routes$route_id)
  
  cta_stop_times <- fread(file.path(cta_dir, "stop_times.txt"))
  rail_stop_ids <- unique(
    cta_stop_times %>% 
    filter(trip_id %in% rail_trips$trip_id) %>% 
    pull(stop_id)
  )
  
  cta_rail_stops <- cta_stops %>% 
    filter(stop_id %in% rail_stop_ids)
}

# Filter Metra stops to Illinois only
metra_il_stops <- metra_stops %>% 
  filter(stop_lat <= 42.5)  # Northern Illinois boundary

# Convert to spatial features
cta_rail_sf <- st_as_sf(
  cta_rail_stops, 
  coords = c("stop_lon", "stop_lat"), 
  crs = WGS84
) %>%
  mutate(agency = "CTA")

metra_sf <- st_as_sf(
  metra_il_stops, 
  coords = c("stop_lon", "stop_lat"), 
  crs = WGS84
) %>%
  mutate(agency = "Metra")

cat(glue("Processed {nrow(cta_rail_sf)} CTA rail stations and {nrow(metra_sf)} Metra stations\n"))
```


```{r create_buffers, include=FALSE}
## Step 3: Create Station Buffers and Load Route Boundaries

# To fix Bug 1, we'll create our own circular buffers instead of relying on pre-buffered boundaries:

cat("Creating station buffers...\n")

# Transform to Illinois State Plane for accurate distance calculations
# Create CTA rail station buffers (2,640 feet)
cta_rail_buffers <- st_transform(cta_rail_sf, IL_STATE_PLANE) %>%
  st_buffer(2640) %>%  # 2,640 feet = 0.5 miles
  st_union() %>%
  st_make_valid()

# Create Metra station buffers (2,640 feet)
metra_buffers <- st_transform(metra_sf, IL_STATE_PLANE) %>%
  st_buffer(2640) %>%
  st_union() %>%
  st_make_valid()

# Load TSL Route boundaries (bus corridors) from Chicago's service
# Layer 14: Route-based TSL (1,320 ft buffers)
cat("Fetching TSL route boundaries...\n")
tsl_route_boundaries <- query_arcgis(14) %>% 
  st_make_valid() %>%
  st_transform(IL_STATE_PLANE)

# Combine all CTA-served areas
cta_served_area <- st_union(cta_rail_buffers, st_union(tsl_route_boundaries))

cat("Buffer creation complete\n")
```


```{r identify_metra_only, include=FALSE}
## Step 4: Identify Metra-Only Areas (Fix for Bug 2)

# Now we'll correctly identify areas that are ONLY served by Metra:

# Metra-only areas are within Metra buffers but NOT within:
# 1. CTA rail station buffers OR
# 2. TSL Route (bus) corridors

# Calculate the correct Metra-only area
metra_only_area <- st_difference(metra_buffers, cta_served_area) %>%
  st_make_valid()

cat("Metra-only area calculation complete\n")
```


```{r apply_carveouts, include=FALSE}
## Step 5: Apply Downtown Carve-out

# Remove downtown districts from the no-parking zone:

# Project and union downtown districts
downtown_union <- st_transform(downtown_districts, IL_STATE_PLANE) %>% 
  st_union() %>%
  st_make_valid()

# CTA areas excluding downtown = no parking requirements
cta_no_parking_zone <- st_difference(cta_served_area, downtown_union) %>%
  st_make_valid()

# All transit-served areas (for statistics)
all_transit_served <- st_union(cta_served_area, metra_buffers) %>%
  st_make_valid()
```


```{r clip_boundaries, include=FALSE}
## Step 6: Clip to City Boundaries

# Ensure all zones are within Chicago city limits:

# Load city boundary
city_boundary <- st_read(
  "https://data.cityofchicago.org/resource/qqq8-j68g.geojson", 
  quiet = TRUE
) %>%
  st_make_valid() %>%
  st_transform(IL_STATE_PLANE)

# Clip all layers to city boundary
cta_no_parking_zone <- st_intersection(cta_no_parking_zone, city_boundary) %>%
  st_make_valid()
metra_only_area <- st_intersection(metra_only_area, city_boundary) %>%
  st_make_valid()
all_transit_served <- st_intersection(all_transit_served, city_boundary) %>%
  st_make_valid()

# Transform back to WGS84 for web mapping
downtown_wgs <- st_transform(downtown_union, WGS84)
cta_no_parking_wgs <- st_transform(cta_no_parking_zone, WGS84)
metra_only_wgs <- st_transform(metra_only_area, WGS84)
all_transit_wgs <- st_transform(all_transit_served, WGS84)
```

```{r calculate_stats, include=FALSE}
# Step 7: Calculate Coverage Statistics
# Let's quantify the impact of this ordinance:

# Helper function to calculate area in square miles
calc_area_sqmi <- function(geometry) {
  area <- st_area(geometry)
  set_units(area, "mi^2") %>% 
    as.numeric() %>% 
    round(1)
}

# Chicago's total area
CHICAGO_AREA_SQMI <- 234

# Calculate areas
total_transit_area <- calc_area_sqmi(all_transit_served)
cta_no_parking_area <- calc_area_sqmi(cta_no_parking_zone)
metra_only_area_sqmi <- calc_area_sqmi(metra_only_area)

# Calculate percentages
pct_total <- round(total_transit_area / CHICAGO_AREA_SQMI * 100, 1)
pct_cta_no_parking <- round(cta_no_parking_area / CHICAGO_AREA_SQMI * 100, 1)
pct_metra_only <- round(metra_only_area_sqmi / CHICAGO_AREA_SQMI * 100, 1)

# Create summary table
coverage_summary <- tibble(
  Zone = c("Total Transit-Served", 
           "No Parking Requirements (CTA)", 
           "Admin Adjustment Required (Metra-only)"),
  `Area (sq mi)` = c(total_transit_area, cta_no_parking_area, metra_only_area_sqmi),
  `% of City` = c(pct_total, pct_cta_no_parking, pct_metra_only)
)

knitr::kable(coverage_summary, caption = "Parking Requirement Zone Coverage")
```

---

## Interactive Map

```{r load_boundaries, include=FALSE}
# Step 8: Interactive Map

# Load community areas for reference lines
cat("Loading community area boundaries...\n")
community_areas <- st_read(
  "https://gisapps.cityofchicago.org/arcgis/rest/services/ExternalApps/extoperational/MapServer/0/query?where=1%3D1&outFields=*&outSR=4326&f=geojson",
  quiet = TRUE
) %>%
  st_make_valid()

# Transform city boundary back to WGS84 for mapping
city_boundary_wgs <- st_transform(city_boundary, WGS84)
```

```{r create_map}
# Create interactive map
parking_map <- leaflet() %>%
  # Base map
  addProviderTiles(providers$CartoDB.Positron) %>%
  
  # Zone polygons
  addPolygons(
    data = cta_no_parking_wgs,
    fillColor = "#6B46C1", 
    fillOpacity = 0.35, 
    color = "#6B46C1", 
    weight = 1,
    group = "No Parking Requirements",
    popup = "No parking requirements - build freely!"
  ) %>%
  
  addPolygons(
    data = metra_only_wgs, 
    fillColor = "#F97316", 
    fillOpacity = 0.45, 
    color = "#EA580C", 
    weight = 2,
    group = "Admin Adjustment Required (Metra-only)",
    popup = "Administrative Adjustment required for parking reduction"
  ) %>%
  
  addPolygons(
    data = downtown_wgs, 
    fillColor = "#000000", 
    fillOpacity = 0.15, 
    color = "#000000", 
    weight = 0.7,
    group = "Downtown Carve-out (D-districts)",
    popup = "Downtown district - parking requirements remain"
  ) %>%
  
  # Transit stations
  addCircleMarkers(
    data = cta_rail_sf, 
    radius = 3, 
    color = "#0078D4", 
    fillOpacity = 0.9, 
    group = "CTA Rail Stations",
    popup = ~paste0("<b>", stop_name, "</b><br>CTA Rail")
  ) %>%
  
  addCircleMarkers(
    data = metra_sf, 
    radius = 3, 
    color = "#E31837", 
    fillOpacity = 0.9, 
    group = "Metra Stations",
    popup = ~paste0("<b>", stop_name, "</b><br>Metra")
  ) %>%
  
  # Reference boundaries (added last to be on top)
  addPolylines(
    data = community_areas,
    color = "#999999",
    weight = 0.5,
    opacity = 0.5,
    fill = FALSE
  ) %>%
  
  addPolylines(
    data = city_boundary_wgs,
    color = "#666666",
    weight = 1,
    opacity = 0.7,
    fill = FALSE
  ) %>%
  
  # Layer controls
  addLayersControl(
    overlayGroups = c("No Parking Requirements", 
                     "Admin Adjustment Required (Metra-only)",
                     "Downtown Carve-out (D-districts)", 
                     "CTA Rail Stations", 
                     "Metra Stations"),
    options = layersControlOptions(collapsed = FALSE)
  ) %>%
  
  # Legend
  addLegend(
    position = "bottomright",
    colors = c("#6B46C1", "#F97316", "#000000"),
    labels = c("No Parking Requirements", 
               "Admin Adjustment Required", 
               "Downtown Carve-out"),
    title = "Parking Elimination Zones",
    opacity = 0.7
  ) %>%
  
  # Hide all groups except the main one
  hideGroup("Admin Adjustment Required (Metra-only)") %>%
  hideGroup("Downtown Carve-out (D-districts)") %>%
  hideGroup("CTA Rail Stations") %>%
  hideGroup("Metra Stations") %>%
  
  # Set initial view
  setView(lng = -87.6298, lat = 41.8781, zoom = 11)

# Display map
parking_map
```

---

## Key Findings

The new ordinance represents a major shift in Chicago's approach to parking requirements:

- **`r total_transit_area` square miles** (`r pct_total`% of the city) are now considered transit-served
- **`r cta_no_parking_area` square miles** (`r pct_cta_no_parking`% of the city) can build residential developments without any parking
- **`r metra_only_area_sqmi` square miles** (`r pct_metra_only`% of the city) require an Administrative Adjustment to reduce parking

This means that approximately **`r pct_cta_no_parking`% of Chicago** is now free from residential parking mandates, enabling more affordable and sustainable development near transit.

---

## Data Sources

- **Transit Data**:
  - [CTA GTFS Feed](https://www.transitchicago.com/developers/gtfs/)
  - [Metra GTFS Feed](https://metra.com/developers)
- **Geographic Boundaries**:
  - [Zoning Districts](https://data.cityofchicago.org/Community-Economic-Development/Boundaries-Zoning-Districts-current-/dj47-wfun)
  - [City Boundary](https://data.cityofchicago.org/Facilities-Geographic-Boundaries/City_Boundary/qqq8-j68g)
  - [TSL Route Boundaries](https://gisapps.chicago.gov/arcgis/rest/services/ExternalApps/Zoning/MapServer/14) (bus corridors)
- **Ordinance**: [O2025-0015577](https://chicago.councilmatic.org/legislation/o2025-0015577/) (Sponsored by Ald. Daniel La Spata, Ald. Andre Vasquez, Ald. Bennett Lawson)

---

## Additional Resources

- [Chicago Cityscape article](https://www.chicagocityscape.com/blog/parking-mandates-2025.php) on the ordinance
- [Chicago Municipal Code - Transit Supportive Bus Line Corridors](https://codelibrary.amlegal.com/codes/chicago/latest/chicago_il/0-0-0-2699368)
- [Department of Planning and Development](https://www.chicago.gov/dpd) for official guidance
- [Connected Communities Ordinance](https://www.chicago.gov/city/en/depts/dcd/supp_info/connected-communities-ordinance.html) background

---

**Questions or feedback?** This analysis is open source. Contributions and corrections are welcome on [Github](https://github.com/MisterClean/chicago-parking-reform/issues).